Crunching Large Graphs with Commodity Processors

### 2011-04-21 doc/hotpar2011/softxmt-hotpar2011.tex
bib updates

\authorinfo{Jacob Nelson$^{\dagger}$,
  Brandon Myers$^{\dagger}$,
  A. H. Hunter$^{\dagger}$,
  Preston Briggs$^{\dagger}$
  Luis Ceze$^{\dagger}$,
  Carl Ebeling$^{\dagger}$,\\
  Dan Grossman$^{\dagger}$,
  Simon Kahan$^{{\dagger \ddagger}}$,
  Mark Oskin$^{\dagger}$,
}{\textdagger University of Washington, \textdaggerdbl Pacific Northwest National Laboratory}{
  \{nelson, bdmyers, ahh, preston, luisceze, ebeling, djg, skahan, oskin\}@cs.washington.edu}

@@

Many important emerging applications such as social network analysis,
bioinformatics, and sensor networks rely on crunching very large
graphs. Unfortunately, the computational cost of these applications
gets quickly out of hand. While tools to analyze social networks and
query semantic graphs with billions of vertices and edges exist
today~\cite{joslyn:hpsf, ediger:msda, brocheler:cosi, pegasus,
  pregel}, graphs of interest to defense applications are expected to
have trillions of entities~\cite{hartley:mssg, kolda:imkd}. Speeding
up these applications at a low cost would have a large impact in how
we analyze large data-sets and make them even more valuable.

@@

Multithreading is a technique that has been used successfully to
implement efficient computations for these
graphs~\cite{bader:bfsmta}. The Cray {\small XMT} is an example of such an
approach: it solves the memory latency problem through concurrency
rather than caching. Each {\small XMT} processor supports 128 hardware contexts
and 1024 outstanding memory operations, and is able to switch contexts
every cycle~\cite{tera, feo-xmt}. This ability comes at a cost: the
{\small XMT} is an expensive, non-commodity machine with low
single-threaded performance.

@@

We observe a gradual decrease in reference rate once the number of
concurrent references per core exceeds 10; we believe this is due to
later prefetches squashing earlier ones in the line fill buffers. In
the multi-node runtime, \texttt{readIssue} will not be implemented
with a prefetch instruction. 
%The single-node coroutine library does
%depend on prefetches: further experiments show that without them, the
%overhead of context switches limits concurrency when there are more
%than two coroutines per core. 
Lastly, we observe repeatable data points that are off the trend and fall below the max. This effect is harder to explain, but we suspect it may be
due to resource contention in the cores' pipelines once the memory
system is full of requests.

@@

% cache pressure figure
\begin{figure}[t]
  \begin{center}
    \includegraphics[width=0.48\textwidth]{figures/cachepressurenewview-edited.pdf}
  \end{center}
  \vspace{-10pt} %space with fig
  \caption{Cache pressure with coroutines, on six cores. The vertical
    axis is remote reference rate, with (a) 8 coroutines per core, and
    (b) 48 coroutines per core, about the number required for the
    network rate in the simulated delay experiments. }
  \vspace{-1.5pt} % space with text
  \label{fig:cache-pressure}
\end{figure}

We varied the size of the per-coroutine local working set from 64~B
to 4~MB, and ran 0 to 32 local updates for each pointer
traversal. We measured the remote reference rate (i.e., accesses to \texttt{remoteList} from Figure~\ref{fig:code}c).

Figure~\ref{fig:cache-pressure}a shows the result for 8~coroutines per
core.  With 1~to 4~updates per remote reference for smaller working
sets, the runtime achieves near maximum throughput. 
%As working set
%size increases, cache misses in the local space become a
%bottleneck. 
%When there are several updates to large working sets, 
As the number of updates and the working set size increase, the data
of other coroutines is likely to be evicted from cache and performance
suffers.

@@

To simulate the performance of pointer chasing in a multi-node system
we referenced the first chip's memory from the second chip's cores and also had our coroutine scheduler include a delay
before a coroutine can be rescheduled, imitating the network transit
delay. We assume 1.1 $\mu$s interface and 400 ns switch latencies in each direction
\cite{infiniband, mellanox:site}, and thus estimate a 3 $\mu$s round-trip
delay. In analyzing the results, we assume a network issue rate of \mrps{100}, based on modern network interfaces \cite{mellanox:press}.

@@

Figure~\ref{fig:cache-pressure}b shows the results of the cache pressure experiment for supporting 48 coroutines per core, about the number of contexts needed to tolerate latency in the network experiment. We observe that when performing up to 8 local updates per remote reference, smaller working sets still allow the desired throughput. When performing 1 to 2 updates, the runtime can maintain throughput for working sets of up to 32~KB.

@@

Much work exists on using multithreading to tolerate latency. Hardware
implementations include the Tera MTA~\cite{tera} and Cray
{\small XMT}~\cite{feo-xmt}, Simultaneous Multithreading \cite{tullsen-smt},
MIT Alewife \cite{agarwal-alewife}, Cyclops \cite{almasi-cyclops}, and
even GPUs \cite{gpus}. Software-only approaches exist as well; the
Software Controlled Multithreading project \cite{mowry-scm}, QThreads
\cite{qthreads}, MAESTRO \cite{maestro}, and Charm \cite{charm} all
use variants of this idea.
%In contrast to all of these, our goal is to tolerate
%latencies across a multi-node system using mostly software.

@@

We presented our plans for building a system for large graph
computation using commodity parts. We leverage decades-old ideas on
using a large amount of concurrency to tolerate latencies, but we do
so mostly in software. We developed a runtime system for latency
tolerance; our results showed that it supports enough concurrency
to tolerate the long latencies of our large high-bandwidth global
memory system.

>>>>

\authorinfo{Jacob Nelson$^{\dagger}$,
  Brandon Myers$^{\dagger}$,
  A. H. Hunter$^{\dagger}$,
  Preston Briggs$^{\dagger}$,
  Luis Ceze$^{\dagger}$,
  Carl Ebeling$^{\dagger}$,\\
  Dan Grossman$^{\dagger}$,
  Simon Kahan$^{{\dagger \ddagger}}$,
  Mark Oskin$^{\dagger}$
}{\textdagger University of Washington, \textdaggerdbl Pacific Northwest National Laboratory}{
  \{nelson, bdmyers, ahh, preston, luisceze, ebeling, djg, skahan, oskin\}@cs.washington.edu}

@@

Many important emerging applications such as social network analysis,
bioinformatics, and sensor networks rely on crunching very large
graphs. Unfortunately, the computational cost of these applications
gets quickly out of hand. While tools to analyze social networks and
query semantic graphs with billions of vertices and edges exist
today~\cite{joslyn:hpsf, ediger:msda, brocheler:cosi, pegasus,
  pregel}, graphs of interest to defense applications are expected to
have trillions of entities~\cite{hartley:mssg, kolda:imkd}. Speeding
up these applications at a low cost would have a significant impact in how
we analyze large data-sets and make them even more valuable.

@@

Multithreading is a technique that has been used successfully to
implement efficient computations for these
graphs~\cite{bader:bfsmta}. The Cray {\small XMT} is an example of such an
approach: it solves the memory latency problem through concurrency
rather than caching. Each {\small XMT} processor supports 128 hardware contexts
and 1024 outstanding memory operations, and is able to switch contexts
every cycle~\cite{tera, feo-xmt}. This ability comes at a cost: the
{\small XMT} is an expensive, non-commodity machine with low
single-thread performance.

@@

We observe a gradual decrease in reference rate once the number of
concurrent references per core exceeds 10; we believe this is due to
later prefetches squashing earlier ones in the line fill buffers. In
the multi-node runtime, \texttt{readIssue} will not be implemented
with a prefetch instruction. 
%The single-node coroutine library does
%depend on prefetches: further experiments show that without them, the
%overhead of context switches limits concurrency when there are more
%than two coroutines per core. 
We also observe repeatable data points that are off the trend and fall below the max. This effect is harder to explain, but we suspect it may be
due to resource contention in the cores' pipelines once the memory
system is full of requests.

@@

% cache pressure figure
\begin{figure}[t]
  \begin{center}
%    \includegraphics[width=0.48\textwidth]{figures/cachepressurenewview-edited.pdf}
    \includegraphics[width=0.48\textwidth]{figures/cachepressure-64to4M-edited2.pdf}
  \end{center}
%  \vspace{-10pt} %space with fig
  \caption{Cache pressure with coroutines, on six cores. The vertical
    axis is remote reference rate, with (a) 8 coroutines per core, and
    (b) 48 coroutines per core, about the number required for the
    network rate in the simulated delay experiments. }
  \vspace{-1.5pt} % space with text
  \label{fig:cache-pressure}
\end{figure}

We varied the size of the per-coroutine local working set from 64~B
to 4~MB, and ran 0 to 32 local updates for each remote pointer
traversal. We measured the remote reference rate (i.e., accesses to \texttt{remoteList} from Figure~\ref{fig:code}c).

Figure~\ref{fig:cache-pressure}a shows the result for 8~coroutines per
core.  With 1~to 4~updates per remote reference for smaller working
sets, the runtime achieves near maximum throughput. 
%As working set
%size increases, cache misses in the local space become a
%bottleneck. 
%When there are several updates to large working sets, 
As the number of updates and the working set size increase, the data
of other coroutines is likely to be evicted from cache, leading to
decreased performance.

@@

To simulate the performance of pointer chasing in a multi-node system
we referenced the first chip's memory from the second chip's cores and also had our coroutine scheduler include a delay
before a coroutine can be reactivated, imitating the network transit
delay. We assume 1.1 $\mu$s interface and 400 ns switch latencies in each direction
\cite{infiniband, mellanox:site}, and thus estimate a 3 $\mu$s round-trip
delay. In analyzing the results, we assume a network issue rate of \mrps{100}, based on modern network interfaces \cite{mellanox:press}.

@@

Figure~\ref{fig:cache-pressure}b shows the results of the cache
pressure experiment for supporting 48 coroutines per core, about the
number of contexts needed to tolerate latency in the network
experiment. We observe that when performing up to 16~local updates per
remote reference, smaller working sets still allow the desired
throughput. When performing 1~to 2~updates, the runtime can maintain
throughput for working sets of up to 32~KB.

@@

Much work exists on using multithreading to tolerate latency. Hardware
implementations include the Tera {\small MTA}~\cite{tera} and Cray
{\small XMT}~\cite{feo-xmt}, Simultaneous Multithreading \cite{tullsen-smt},
MIT Alewife \cite{agarwal-alewife}, Cyclops \cite{almasi-cyclops}, and
even GPUs \cite{gpus}. Software-only approaches exist as well; the
Software Controlled Multithreading project \cite{mowry-scm}, QThreads
\cite{qthreads}, MAESTRO \cite{maestro}, and Charm \cite{charm} all
use variants of this idea.
%In contrast to all of these, our goal is to tolerate
%latencies across a multi-node system using mostly software.

@@

We presented our plans for building a system for large graph
computation using commodity parts. We leverage decades-old ideas on
using a large amount of concurrency to tolerate latencies, but we do
so mostly in software. We developed a runtime system for latency
tolerance, and our results showed that it supports enough concurrency
to tolerate the long latencies of our large high-bandwidth global
memory system.
